---
title: "crossrun. An R Package for the Joint Distribution of Number of Crossings and  
  Longest Run in Independent Bernoulli Observations"
shorttitle: "crossrun"
author: 
- name: Tore Wentzel-Larsen  
  affiliation: 
    - Centre for Child and Adolescent Mental Health, Eastern and Southern Norway;
     Norwegian Centre of Violence and Traumatic Stress Studies 
  email: tore.wentzellarsen@gmail.com
- name: Jacob Anhøj
  affiliation: 
  - Rigshospitalet, University of Copenhagen, Denmark
  email: jacob@anhoej.net
package name: crossrun
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=7.15,
  fig.height=3.5)
```


## Introduction

The setting is defined by a number n of independent observations from a Bernoulli distribution with the same success probability. In statistical process control, our main intended application, this may be the useful observations in a runs plot, recording values above and below a pre-specified median, disregarding any observations equal to the median ([Anhøj (2015)](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0121349)). The focus of the R package `crossrun` is the number C of crossings and the length L of the longest run in such a sequence. A run is a sequence of successes or failures, delimited by a different observation or the start or end of the entire sequence. A crossing is two adjacent different observations. The following figure illustrates runs and crossings in a statistical process control setting, with successes and failures represented by observations above and below the midline, respectively, in n=24 observations:

```{r preparefig1, echo=FALSE, results="hide", message=FALSE}
set.seed(32)
y <- rnorm(24)
x <- seq_along(y)
```

```{r fig1, echo=FALSE, message=FALSE, fig.cap="Figure 1"}
op <- par(mar = c(bottom = 3,
                  left = 0,
                  top = 0,
                  right = 0
))

plot(x, y,
     axes = FALSE,
     type = "b",
     pch = '',
     lwd = 1.5,
     xlab = "",
     ylab = ""
)

lines(x, rep(median(y), 24),
      col = 'grey40')

text(1:24, y)

par(op)
```

In Figure 1 the longest run consists observations 12-16 below the midline. The second longest run, above the midline, consists of observation 4-7. Thus the length L of the longest run is 5 in this case. The number of crossings of the midline is C=11. 

What is computed in the package `crossrun` is the joint distribution of C and L. An article describing the rationale behind these computations is planned and will be announced at the GitHub site for `crossrun` (https://github.com/ToreWentzel-Larsen/crossrun). L and C are integers. L may be any integer between 1, if all subsequent observations are different, and n, if all observations are equal. C may be any integer between 0, if all observations are equal, and n-1 if all subsequent observations are different. Not all combinations of C and L are possible as shown in the following example for n=15 and success probability 1/2:

```{r symm15prep, echo=FALSE, results="hide", message=FALSE}
library(crossrun)
j15s <- joint15symm
rownames(j15s)[1] <- "C=0"
colnames(j15s)[1] <- "L=1"
```

```{r symm15, echo=FALSE, message=FALSE}
knitr::kable(j15s)
```

As will be described and justified in more detail later, the table above does not give the probabilities themselves, but the probabilities multiplied by $2^{n-1}$ that is 16384 for n=15. For instance $P (C=6, L=5)$ = 861/16384 = 0.053. The highest joint probability is $P (C=8, L=3)$ = 1470/16384 = 0.090. As seen in the table the non-zero probabilities are confined to a sloped region that is rather narrow, but sufficiently wide that the two variables together are more informative than each of them in isolation. These are general phenomona. 

The procedure for computing the joint distribution of C and L is iterative, this means that the joint distribution for a sequence of length n cannot be computed before the joint distributions for all shorter sequences have been computed. At the moment the computations have been performed and checked for n up to 100, and success probabilities 0.5 to 0.9 in steps of 0.1.

## The main function: `crossrunbin`

The function `crossrunbin` has two main arguments, `nmax` that is the maximum sequence length and `prob` that is the success probability. Other arguments include a multiplier described later and a precision parameter, these should normally be left at their default values, and `printn` that allows progress information during a usually lengthy computation. For details see 

`?crossrunbin`

The joint probabilities are stored as 6 lists of which `pt` is sufficient for normal use, the others are mainly included for code checking. `pt`, itself a list, gives an n by n matrix (well, not quite, but something closely related as detailed below) for each sequence length n. For instance

`crb40.6 <- crossrunbin(nmax=40, prob=.6, printn=TRUE)$pt`

computes the joint probabilities for all sequence lenghts $n \leq 40$ when the success probability is 0.6. For simplicity, the command above only returns the joint probabilities `pt`. When the computation is finished, the joint distribution for say n=15 is 

`crb40.6[[15]]`

Actually the resulting joint distribution is not quite a matrix, it is a two-dimensional mpfr array ([Fousse et, al, 2007](http://doi.acm.org/10.1145/1236463.1236468), [Mächler 2018](https://CRAN.R-project.org/package=Rmpfr)). Two-dimensional mpfr arrays are almost the same as matrices, but with appreciably higher precision. Since the computation procedure is iterative high precision during calculation is vital, but the resulting joint distributions may subsequently be transformed into ordinary matrices by the [Rmpfr](https://CRAN.R-project.org/package=Rmpfr) function `asNumeric` for easier presentation. To limit within package storage only the joint distributions for n=15, 60, 100 and success probabilities 0.5 (the symmetric case) and 0.6 have been included in the package, as ordinary matrices.

## The "times" representation of the joint distributions

The joint distributions are actually not computed as probabilities, but as probabilities multiplied by a multiplier whose default value is $2^{n-1}$. Optionally another multiplier $m^{n-1}$ could be used where $m$ is an argument (`mult`) to the function `crossrunbin`, but the default value should normally be used. This representation is shown above for n=15 in the symmetric case (p=1/2). One may note that in the example above all probabilities are represented by integers in the "times" representation. This is a general phenomenon in the symmetric case, but not for success probabilites different from 1/2. In the symmetric case ([Anhøj and Vingaard Olesen (2014)](http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0113825)) the number C of crossings has a binomial distribution with number of observations n-1 and probability 1/2, and their marginal probabilities are just the binomial coefficients divided by $2^{n-1}$. Indeed, the row sums in the times representation are the binomial coefficients in the symmetric case. This will be explicated later in the case n=15.

When the success probability is not 1/2 the joint distribution is no longer represented by integers even in the times representation. This is illustrated below for n=15 and success probability 0.6. 

```{r p6.15prep, echo=FALSE, results="hide", message=FALSE}
library(crossrun)
p6.15 <- joint15.6
rownames(p6.15)[1] <- "C=0"
colnames(p6.15)[1] <- "L=1"
```

```{r p6.15, echo=FALSE, message=FALSE,fig.width=15}
knitr::kable(round(p6.15,1))
```

In the table above the probabilites are shown with one decimal, and in the times repesentation (multiplied by $2^{n-1}$=16384). The same probabilities are different from 0 as in the symmetric case, but the distribution has been shifted in the direction of longer runs and fewer crossings. 

The times representation may be advantageous for presention since very small numbers are avoided. The main reason for using this representation is, however, to enhance precision in the iterative computation procedure.

## The symmetric case: `crossrunsymm`

In the symmetric case the joint probabilities are, as illustrated above, stored as integers in the times representation. A separate function `crossrunsymm` is available in this case. The arguments, except the success probability, are the same as in the more general function `crossrunbin`. The inner workings are somewhat simpler as will be accounted for in more detail in the forthcoming article. 

## Generalization

The procedure cannot be generalized to autocorrelated time sequences. Also, in statistical process control the procedure does only apply when the median is pre-specified, not when it is determined by the time series itself. However, in the case with non-constant success probability a similar procedure is available and implemented in a function `crossrunchange`. In this procedure all arguments are as in `crossrunbin`, except that the success probability is replaced by a vector of length n with success probabilities for each of the n time points. 

## Checking procedures

Procedures for checking the joint distributions are available in `crossrun`. First, the function `exactbin` computes the joint distribution for $n \leq 6$ independently of the iterative procedure, by formulas based on "brute force" enumaration that is practically feasible for these short sequences. An example of use of `exactbin` for checking of results of the exact procedure is as follows, for n=5 and success probability 0.6 (multiplied by $2^4=16$ to conform with the times representation):

```{r exact1, message=FALSE}
library(crossrun)
library(Rmpfr)
exact1 <- asNumeric((2^4)*exactbin(n=5, p=0.6))
iter1 <- asNumeric(crossrunbin(nmax=5, prob=.6)$pt[[5]])
compare1 <- cbind(exact1,iter1)
knitr::kable(compare1)
```

Here the 5 leftmost columns come from exact calculations while the 5 rightmost columns come from the iterative procedure. The maximum absolute difference is computed as `r max(abs(exact1-iter1))` in this case. Generally some tiny differences may occur.

The iterative computations may also be checked for appreciably higher n. As commented above the row sums in the symmetric case are just the binomial coefficients. The following code checks this fact for n=15.

```{r bincoeff15, message=FALSE}
library(crossrun)
library(Rmpfr)
bincoeff14 <- Rmpfr::chooseMpfr.all(14) # binomial coefficients, n-1=14
bincoeff14iter <- cumsumm(j15s)[-1,15] # row sums, n-1=14
compare15 <- rbind(asNumeric(bincoeff14), bincoeff14iter)
row.names(compare15) <- c("exact","iter")
compare15
max(abs(bincoeff14 - bincoeff14iter))
```
This check may be repeated by the same procedure for appreciably higher n, but then full mpfr precision should be used for the results of the iterative procedure. Then no discrepancy has been found for $n \leq 100$.

The results of the iterative procedure may also be checked by results of simulations. The `crossrun` function `simclbin` performs simulations for the number of crossings and the longest run for chosen values of the success probability. Again, computations for substantially longer sequences (n appreciably higher than 15) should use full mpfr precision for the joint distribution. The following code shows the procedure for n=15 and 10000 simulations and compares the mean of $C \cdot L$ in the simulations with the corresponding mean from the joint distribution p6.15 with success probability 0.6:

```{r sim15, message=FALSE}
library(crossrun)
set.seed(83938487)
sim15 <- simclbin(nser=15, nsim=10000)
(matrix(0:14,nrow=1) %*% p6.15 %*% matrix(1:15,ncol=1))/2^14
mean(sim15$nc0.6*sim15$lr0.6)
```

Here, $C \cdot L$ is just used as an example of a fairly demanding function of the number C of crossing and the longest run L. Again, computations for substantially longer sequences (n appreciably higher than 15) should use full mpfr precision for the joint distribution. Simpler statistics include means and standard deviations of C and L separately. The following shows a graphical comparison of the cumulative distribution functions of C and L based on the joint distribution and the simulations. 

```{r sim15plot, message=FALSE}
library(crossrun)
plot(x=as.numeric(names(table(sim15$nc0.6))),
     y=(cumsum(cumsumm(p6.15)[,15])/(2^14))[
       as.numeric(names(table(sim15$nc0.6)))+1], 
     type="l", xlab="Number of crossings", ylab="CDF", las=1)
points(x=as.numeric(names(table(sim15$nc0.6))), 
       y=cumsum(table(sim15$nc0.6))/sum(table(sim15$nc0.6)),
       type="l", col="red",lty="dotted")
lines(x=c(0,1), y=c(.9,.9), col="red")
text(x=1, y=.9, pos=4, labels="red: simulations", col="red")
plot(x=as.numeric(names(table(sim15$lr0.6))), 
     y=as.numeric(cumsum(cumsummcol(p6.15)[15,])/sum(cumsummcol(p6.15)[15,]))[
       as.numeric(names(table(sim15$lr0.6)))],
     type="l", xlab="Longest run", ylab="CDF", las=1)
points(x=as.numeric(names(table(sim15$lr0.6))), 
       y=cumsum(table(sim15$lr0.6))/sum(table(sim15$lr0.6)),
       type="l", col="red", lty="dotted")
lines(x=c(2,3), y=c(.9,.9), col="red")
text(x=3, y=.9, pos=4, labels="red: simulations", col="red")
```

## References

1. Jacob Anhøj (2015). [Diagnostic value of run chart analysis: 
 Using likelihood ratios to compare run chart rules on simulated 
 data series](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0121349)
PLOS ONE 10 (3): e0121349.

1. Jacob Anhøj, Anne Vingaard Olesen (2014). [Run Charts Revisited: A Simulation Study of Run Chart Rules for Detection of Non-Random Variation in Health Care Processes](http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0113825). PLoS ONE 9(11): e113825.
 
1. Laurent Fousse, Guillaume Hanrot, Vincent Lefèvre, 
 Patrick Pélissier, Paul Zimmermann. (2007). [Mpfr: A multiple-precision 
 binary floating-point library with correct 
 rounding](http://doi.acm.org/10.1145/1236463.1236468)
ACM Trans. Math. Softw. 33 (2): 13. ISSN 0098-3500.

1. Martin Mächler (2018). [Rmpfr: R MPFR - Multiple Precision 
  Floating-Point Reliable](https://CRAN.R-project.org/package=Rmpfr)
R package version 0.7-0.


